---
title: "Insights into the dynamics of wolf occupancy in human-dominated landscapes"
author: "Mónia Nakamura, José Vicente López-Bao, Helena Rio-Maior, Sara Roque, Patrícia
  Gil, Ana Serronha, Emilio García, Orencio Hernández Palacios, Gonçalo Ferrão da
  Costa, Francisco Álvares, Francisco Petrucci-Fonseca, Olivier Gimenez, Pedro Monterroso"
subtitle: "R code - Dynamic occupancy model for wolf in the Iberian Peninsula"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_knit$set(eval = FALSE,
                     echo=TRUE,
                     message = FALSE,
                     warning = FALSE,
                     results = 'hide',
                     dpi = 300)
```

# Load required packages

```{r}
library(R2jags)
```

# Load data

```{r}
load('D:/PhD/TASK_3_Occupancy/R/OccupancyData/dataMN/RData_files/data_cov_umf_2005_2022.RData') # REMOVE AFTER
# load('Data/data_cov_umf_2005_2022.RData')  
```

# Format detection/nondetection data in a 3D array

```{r}
yy <- data_cov_umf_2005_2022@y
dim(yy)
( nsites <- dim(yy)[1] )
( nsurveys <- 4)
( nyears <- dim(yy)[2]/nsurveys )
y <- array(NA, dim = c(nsites, nsurveys, nyears))
for(i in 1:nyears){
  #print((4*i-3):(4*i))
  y[,,i] <- yy[,(4*i-3):(4*i)]
}
```

# Observation Covariates (Detection): grab rugged, density track, and sampling effort 

```{r}
rugged <- data_cov_umf_2005_2022@siteCovs$rugged  # Ruggedness

densityTrack <- data_cov_umf_2005_2022@siteCovs$densityTrack  # Unpaved road density

effortt <- matrix(data_cov_umf_2005_2022@obsCovs$effort, nrow = 395, ncol = 72, byrow = TRUE)
effort <- array(NA, dim = c(nsites, nsurveys, nyears))
for(i in 1:nyears){
  #print((4*i-3):(4*i))
  effort[,,i] <- effortt[,(4*i-3):(4*i)]
}
binary.effort <- effort
binary.effort[is.na(binary.effort)] <- 0
binary.effort[binary.effort!=0] <- 1

effort[is.na(effort)] <- 0   # Transect effort
```

Site covariates and Yearly site covariates 
```{r}
# Site covariates
densityHigh <- data_cov_umf_2005_2022@siteCovs$densityHigh              # Highway density
densityNatioRegio <- data_cov_umf_2005_2022@siteCovs$densityNatioRegio  # National-Regional road density
densityLocal <- data_cov_umf_2005_2022@siteCovs$densityLocal            # Local road density
densityTrack                                                            # Unpaved road density
alt <- data_cov_umf_2005_2022@siteCovs$alt                              # Altitude
LU_den <- data_cov_umf_2005_2022@siteCovs$LU_den                        # Livestock Unit Density

# Bundle site covariates for extinction parameter
Xeps <- as.matrix(cbind(densityHigh, densityNatioRegio, densityLocal, densityTrack, alt, LU_den))
# Bundle site covariates for colonization parameter
Xgam <- as.matrix(cbind(densityHigh, densityNatioRegio, densityLocal, densityTrack, alt, LU_den)) 

# Yearly site covariates
agric <- matrix(data_cov_umf_2005_2022@yearlySiteCovs$agric, nrow = nsites, ncol = nyears, byrow = T)           # Proportion of Agricultural Land
burn <- matrix(data_cov_umf_2005_2022@yearlySiteCovs$burn, nrow = nsites, ncol = nyears, byrow = T)             # Proportion of Burned Areas
gpw_popden <- matrix(data_cov_umf_2005_2022@yearlySiteCovs$gpw_popden, nrow = nsites, ncol = nyears, byrow = T) # Human Population density
```

## Fit dynamic occupancy model in Jags

# Bundle and summmarize data set
```{r}
str(bdata <- list(y = y, 
                  nsites = nsites,
                  nsurveys = nsurveys, 
                  nyears = nyears,
                  npareps = ncol(Xeps),
                  npargam = ncol(Xgam),
                  Xeps = Xeps,
                  Xgam = Xgam,
                  rugged = rugged,
                  effort = effort,
                  densityTrack = densityTrack,
                  beffort = binary.effort,
                  agric = agric,
                  burn = burn,
                  popden = gpw_popden))
```

# Specify model in BUGS language
```{r penalizing priors adding variables}
pendynamic <- function() {
  # Priors
    # Initial Occupancy
  for(i in 1:nsites) {      
    thetapsi[i] ~ dnorm(0, tau.thetapsi)
  }
    tau.thetapsi <- 1/ (sd.thetapsi * sd.thetapsi)
    sd.thetapsi ~ dunif(0, 5)
    alphapsi ~ dnorm(0,0.1) # Initial occupancy

   # Persistence
  for(i in 1:nsites) {      
    thetaeps[i] ~ dnorm(0, tau.thetaeps)
  }
    tau.thetaeps <- 1/ (sd.thetaeps * sd.thetaeps)
    sd.thetaeps ~ dunif(0, 5)
    alphaeps ~ dnorm(0,0.1) 
  
  # L1 regularization == a Laplace (double exponential) prior 
  for (ss in 1:npareps){
    betaeps[ss] ~ ddexp(0, 1/lambdaeps)
  }
  lambdaeps ~ dt(0, 1, 1) # Cauchy(0,1)
  
    # Colonization
  for(i in 1:nsites) {      
    thetagam[i] ~ dnorm(0, tau.thetagam)
  }
    # mu.thetagam ~ dnorm(0, 0.01)
    tau.thetagam <- 1/ (sd.thetagam * sd.thetagam)
    sd.thetagam ~ dunif(0, 5)
    alphagam ~ dnorm(0, 0.1) 
  
  # L1 regularization == a Laplace (double exponential) prior 
  for (gg in 1:npargam){
    betagam[gg] ~ ddexp(0, 1/lambdagam)
  }
  lambdagam ~ dt(0, 1, 1) # Cauchy(0,1)
  
  # site*year covariates 
  for (jj in 1:3){      ## 4 covariates (agric, burn, popden) 
    betaepsyear[jj] ~ ddexp(0, 1/lambdaeps)
    betagamyear[jj] ~ ddexp(0, 1/lambdagam)
  }
  
  # Detection
  for(i in 1:nsites) {      
    thetap[i] ~ dnorm(0, tau.thetap)
  }
    # mu.thetap ~ dnorm(0, 0.01)
    tau.thetap <- 1/ (sd.thetap * sd.thetap)
    sd.thetap ~ dunif(0, 5)
    alphap ~ dnorm(0, 0.1)
    
  betap[1] ~ dnorm(0, 0.1)  
  betap[2] ~ dnorm(0, 0.1)
  betap[3] ~ dnorm(0, 0.1) # densityTrack

  # Likelihood
  # Ecological submodel
  for (i in 1:nsites){
    z[i,1] ~ dbern(psi[i])
    logit(psi[i]) <-  alphapsi + thetapsi[i]  
    
    for (t in 2:nyears){
      z[i,t] ~ dbern(z[i,t-1] * (1-epsilon[i,t-1]) + (1-z[i,t-1]) * gamma[i,t-1])
      
      logit(epsilon[i,t-1]) <- alphaeps +  thetaeps[i] +  inprod(Xeps[i,], betaeps) +
        betaepsyear[1] * agric[i,t-1] + 
        betaepsyear[2] * burn[i,t-1] +  betaepsyear[3] * popden[i,t-1]
      
      logit(gamma[i,t-1]) <-  alphagam + thetagam[i] + inprod(Xgam[i,], betagam) +  
        betagamyear[1] * agric[i,t-1] + 
        betagamyear[2] * burn[i,t-1]  + betagamyear[3] * popden[i,t-1]
    }
  }
  
  # Observation submodel
  for (i in 1:nsites){
    for (j in 1:nsurveys){
      for (t in 1:nyears){
        p[i,j,t] <- (1 / (1 + exp(-lp[i,j,t]))) * (1-step(-beffort[i,j,t]))
        lp[i,j,t] <- alphap +  thetap[i] + betap[1] * effort[i,j,t] + betap[2] * rugged[i] + betap[3] * densityTrack[i]
       
        y[i,j,t] ~ dbern(z[i,t] * p[i,j,t])
       }
      }
    }
  }
```

# Initial values
```{r}
zst <- array(1, dim = c(nsites, nyears)) # Cheap inits for z
inits <- function(){ list(z = zst, lambdaeps = .1, lambdagam = .1,
                          thetapsi = runif(395, -1, 1),
                          thetaeps = runif(395, -1, 1),
                          thetagam = runif(395, -1, 1),
                          thetap = runif(395, -1, 1))
}
```

# Parameters monitored
```{r}
params <- c ("thetapsi", "sd.thetapsi",
             "betaeps", "thetaeps", "sd.thetaeps",
             "betagam", "thetagam", "sd.thetagam",
             "betap", "thetap", "sd.thetap",
            "lambdaeps", "lambdagam", 
            "betaepsyear", "betagamyear", 
            "alphapsi", "alphaeps",  "alphagam", "alphap",
            "z")
```

# MCMC settings
```{r}
ni <- 40000
nt <- 1
nb <- 3000
nc <- 2  
```

# Call JAGS, check convergence and summarize posteriors
```{r}
model_output <- jags(
  data = bdata, 
  inits = inits, 
  parameters.to.save = params, 
  model.file = pendynamic, 
  n.chains = nc, 
  n.thin = nt, 
  n.iter = ni, 
  n.burnin = nb)
```